Combined annotation exploration for SCPCP000006

### 2024-10-30 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
changes to PR844

---
title: "Annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.85
  cnv_threshold: 0
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

_Where `cnv.thr` and `map.thr` need to be discussed_

first   level annotation | second level   annotation | selection of   the cells | marker genes   for validation | cnv validation
-- | -- | -- | -- | --
normal | endothelial | compartment == "endothelium"     & mapping_score > map.thr     & cnv_score < cnv.thr | WVF | no   cnv
normal | immune | compartment == "immune"     & mapping_score > map.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & mapping_score > map.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
normal | stroma | compartment == "stroma"     & mapping_score > map.thr     & cnv_score < cnv.thr | VIM | no cnv
cancer | stroma | compartment == "stroma"     & cnv_score > cnv.thr | VIM | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | blastema | compartment ==   "fetal_nephron"     & cell_type == "mesenchymal cell"     & cnv_score > cnv.thr | CITED1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | epithelial | compartment ==   "fetal_nephron"     & cell_type != "mesenchymal cell"     & cnv_score > cnv.thr | CDH1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
unknown | - | the rest of the cells | - | proportion_cnv_chr     -1     -4     -11     -16     -17     -18

@@

#### do_Feature_boxplot

`do_Feature_boxplot` shows boxplot of expression of a feature grouped by a metadata.

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_boxplot <- function(df, group.by, feature, split.by ){

  df <- df %>%
    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))

  p <- ggplot(df, 
    aes(x = group, y = !!sym(feature), fill = group)
  ) + 
  geom_boxplot(size = 0.5, size.outlier = 0.25) +
  facet_wrap(vars(!!sym(split.by)), scale = "free_y", ncol = 4) + 
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0), 
    legend.position = "none"
  )

@@

#### CNV score
  cell_type_df$has_cnv_score <- rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])
  cell_type_df$has_cnv_score[cell_type_df$has_cnv_score > params$cnv_threshold] <- "CNV"

  # we know that some immune cells might have false positive CNV 
  cell_type_df$has_cnv_score[cell_type_df$has_cnv_score %in% c(0:params$cnv_threshold)] <- "no CNV"

@@

cell_type_df$first.level_annotation <- "unknown"

# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") && has_cnv_score == "no CNV" ~ "normal",
    compartment %in% c("endothelium", "immune") && has_cnv_score == "no CNV" ~ "normal",
    compartment %in% c("fetal_nephron", "stroma") && has_cnv_score == "CNV" ~ "cancer",
    # all other cells are unknown
    .default ~ "unknown"
  ))

@@

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "kidney"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("stroma") &
                                      cell_type_df$has_cnv_score == "no CNV"]                          <- "normal stroma"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("endothelium") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "endothelium"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("immune") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV") ]                        <- "immune"

@@

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("stroma") &
                                      cell_type_df$has_cnv_score == "CNV"]                          <- "cancer stroma"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "CNV" &
                                        cell_type_df$cell_type == "mesenchymal cell"]                          <- "blastema"

cell_type_df$second.level_annotation[  cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "CNV" &
                                        cell_type_df$cell_type != "mesenchymal cell"]                          <- "cancer epithelium"

>>>>

---
title: "Combined annotation exploration for SCPCP000006"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  predicted.celltype.threshold: 0.85
  cnv_threshold: 0
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

_Where `cnv.thr` and `pred.thr` need to be discussed_

first   level annotation | second level   annotation | selection of   the cells | marker genes   for validation | cnv validation
-- | -- | -- | -- | --
normal | endothelial | compartment == "endothelium"     & predicted.score > pred.thr     & cnv_score < cnv.thr | WVF | no   cnv
normal | immune | compartment == "immune"     & predicted.score > pred.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & predicted.score > pred.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
normal | stroma | compartment == "stroma"     & predicted.score > pred.thr     & cnv_score < cnv.thr | VIM | no cnv
cancer | stroma | compartment == "stroma"     & cnv_score > cnv.thr | VIM | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | blastema | compartment ==   "fetal_nephron"     & cell_type == "mesenchymal cell"     & cnv_score > cnv.thr | CITED1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | epithelial | compartment ==   "fetal_nephron"     & cell_type != "mesenchymal cell"     & cnv_score > cnv.thr | CDH1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
unknown | - | the rest of the cells | - | proportion_cnv_chr     -1     -4     -11     -16     -17     -18

@@

@@

cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold ~ "CNV",
    has_cnv_score <= params$cnv_threshold ~ "no CNV"
  ))

@@

# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "no CNV" ~ "normal",
    compartment %in% c("endothelium", "immune") & cell_type.score > params$predicted.celltype.threshold ~ "normal",
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "CNV" ~ "cancer",
    .default = "unknown"

  ))

@@

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
   cell_type_df$compartment %in% c("fetal_nephron") &
                                      cell_type_df$has_cnv_score == "no CNV"                          ~ "kidney", 
   cell_type_df$compartment %in% c("stroma") &
                                      cell_type_df$has_cnv_score == "no CNV"                          ~ "normal stroma",
   cell_type_df$compartment %in% c("endothelium") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV")                         ~ "endothelium",
   cell_type_df$compartment %in% c("immune") &
                                      (cell_type_df$compartment.score > params$predicted.celltype.threshold |
                                      cell_type_df$has_cnv_score == "no CNV")                         ~ "immune",

   # assign cancer cells based on condition
   cell_type_df$compartment %in% c("stroma") &
          cell_type_df$has_cnv_score == "CNV"                         ~ "cancer stroma",

          cell_type_df$compartment %in% c("fetal_nephron") &
          cell_type_df$has_cnv_score == "CNV" &
          cell_type_df$cell_type == "mesenchymal cell"                ~ "blastema",

          cell_type_df$compartment %in% c("fetal_nephron") &
          cell_type_df$has_cnv_score == "CNV" &
          cell_type_df$cell_type != "mesenchymal cell"                ~ "cancer epithelium",
    .default = "unknown"

  ))
